Efficient pedigree recording for fast population genetics simulation

### 2017-11-24 forwards_paper.tex
merged changes from jeromekelleher/edit-pass
merged methods and results; put benchmarks first

previous \textbf{OUTLINE}, possibly out-of-date but left for comparison.
\begin{enumerate}
    \item motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
    \item explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
    \item review something about msprime methods for storing/traversing tree sequence
    \item describe tables and write out conditions to have valid tables
    \item write down algorithm used to do simple WF simulation
    \item describe simplify algorithm
    \item back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
    \item comparison of speed with simupop, fwdpp
\end{enumerate}

@@

Below, we first describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{})
to efficiently record simplified a tree sequence,
and analyze of its run time complexity.
This is followed by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
to the tools in \msprime{}.

@@

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence,
and each node refers to a distinct ancestor.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with.
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born at time 1 and 2 respectively.

@@

@@

\jk{Insert lost text about simplify efficiency and plot}

\subsection*{Forward simulation and sequential simplification}

\subsection*{Simulation benchmarks}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}

\plr{some of this can move back to an appendix, but since they want methods at the end,
it's the sort of thing they want here.}

%%%%%%%%%%%%%%
\subsection*{Simuation methods}

Jaime and Kevin to describe simulation methods here (separately).

%%%%%%%%%%%%%%
\subsection*{The table format}

Given a set of node and edge tables as described above,
there are a small set of requirements that ensure the tables
describe a valid tree sequence.
(There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guarenteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
\jda{TODO add function name}

For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
and then using \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

Note that since each node time is equal to the amount of time since the \emph{birth} of the
corresponding parent, time is measured in clock time, not in meioses.
\jda{is this the right place? I'd put it in discussion of Figure~\ref{fig:example_tree_sequence}}

%%%%%%%%%%%%%%
\subsection*{Simplification}

@@

>>>>

@@

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{})
to efficiently record simplified a tree sequence,
and analyze of its run time complexity.

%%%%%%
\subsection*{Simulation benchmarks}

\plr{Jaime and Kevin to describe simulation methods here (separately).}

@@

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence,
and each node refers to a distinct ancestor.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with.
Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born at time 1 and 2 respectively.

@@

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are a small set of requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guarenteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
\jda{TODO add function name}
For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
and then using \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

@@

More concretely,
the algorithm works by moving back through time,
processing each parent in the input tree sequence in chronological order.
The main state of the algorithm at each point in time is a set of ancestral lineages,
and each lineage is a collection of ancestral segments.
An ancestral segment $(\ell, r, u)$ is found in a lineage
if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
% These segments are stored in a collection of linked lists,
% one list of segments for each ancestral lineage present at that time.
(These lineages are the ``colors'' above.)
We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.
Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.

%%%%%%%%%%%%%%
\subsection*{Simplification algorithm}

@@

%%%%%%%%%%%%%%
\subsection*{Forward simulation and sequential simplification}

%%%%%%%%%%%%%%
% \section*{Methods}
